---
title: "Exercise_GLM"
author: "Hanh & Verena"
format: html
editor: visual
toc: true
toc-depth: 4
number-sections: true
editor_options: 
  chunk_output_type: console
---

# Exercise for lecture 3 - LMM {.unnumbered}

The main reference, where we also got the data from are:

-   Zuur et al. (2017) Spatial, temporal and spatial-temporal ecological data analysis with R-INLA. Volume I: Using GLM and GLMM
-   <https:://www.highstat.com>

# Load packages

```{r}
library(lattice)
library(ggplot2)
library(dplyr)
library(lme4)     # for function lmer() 
library(nlme)     # for function lme()
```

# Exercise - Linear mixed effect model

Load data

```{r}
WS <- read.csv(file = "White_Stork_Growth_20112012_V2.csv", header = TRUE)
```

Data source within Zuur et al. (2017): Bouriach et al., 2015 
Study of factors affecting growth of White Stock (Ciconia ciconia) nestlings in eastern Algeria

## Data analysis and extract

Data summary

```{r}
str(WS)
summary(WS)
```

Select the 2012 data

```{r}
WS1 <- subset(WS, Year == 2012)
```

Remove NA

```{r}
WS2 <- na.exclude(WS1[, c("Beak", "Age", "Nest", "Chick")])
```

Factor variables

```{r}
WS2$fNest  <- factor(WS2$Nest)
WS2$fChick <- factor(WS2$Chick)
```

Check frequency of data

```{r}
table(WS2$fNest)
table(WS2$fChick)
```

# Plot growth of beak length (mm) versus chick age (days)

```{r}
ggplot(data = WS2, aes(x = Age, y = Beak)) +
  geom_point(shape = 1, size = 1) +
  labs(x = "Age (days)", y = "Beak length (mm)") +
  theme(text = element_text(size = 12))
```

# Models

## Model 1 - Wrong model: fNest as a random effect (ignoring random effects of fChick)

create same model with two different packages / functions

```{r}
MM1_lmer <- lmer(Beak ~ Age + (1 | fNest), data = WS2)
summary(MM1_lmer)

MM1_lme <- lme(Beak ~ Age, random = ~ 1 | fNest, data = WS2)
summary(MM1_lme)
```

Validation plots (only for lme)

```{r}
fit <- fitted(MM1_lme)
res <- residuals(MM1_lme)

par(mfrow = c(2, 2))
plot(fit, res, main = "Residuals vs Fitted")
abline(h = 0, col = "red")
qqnorm(res)
qqline(res, col = "red")
hist(res, breaks = 30, main = "Histogram of Residuals")
```

### Plot simulated beaker lenght versus Age using MM1_lme model

prepare data to create the plot, claculate SE

```{r}
MyData <- data.frame(Age = seq(min(WS2$Age), 
                          max(WS2$Age),
                          length = 25))
X <- model.matrix(~ Age, data = MyData)
MyData$Pred <- X %*% fixef(MM1_lme)
MyData$SE <- sqrt(diag(X %*% vcov(MM1_lme) %*% t(X))  )
MyData$SeUp <- MyData$Pred + 1.96 * MyData$SE
MyData$SeLo <- MyData$Pred - 1.96 * MyData$SE
```

create plot

```{r}
ggplot() + 
  geom_point(data = WS2, aes(x = Age, y = Beak), shape = 1, size = 1) +
  geom_line(data = MyData, aes(x = Age, y = Pred), colour = "black") +
  geom_ribbon(data = MyData, aes(x = Age, ymin = SeLo, ymax = SeUp), alpha = 0.6) + # add ribbin of SE
  labs(x = "Age (days)",
    y = "Beak length (mm)") +
  theme(text = element_text(size = 12))
```

### Plot with fixed and random effects

create plot where each line is one chick

```{r}
WS2$Fit <- fitted(MM1_lme)
ggplot(WS2, aes(x = Age, y = Beak)) +
  geom_point(shape = 1, size = 1) +
  geom_line(aes(y = Fit, group = fNest, color = fNest), show.legend = FALSE) +
  labs(x = "Age (days)", y = "Beak length (mm)") +
  theme(text = element_text(size = 12))
```

## Model 2 - Correct model: Two-way nested model with fNest and fChick

create model with lmer

```{r}
MM2_lmer <- lmer(Beak ~ Age + (1 | fNest / fChick), data = WS2)
summary(MM2_lmer)
```

create model with lme, interaction of nest and chick need to be added manually

```{r}
WS2$fNest_fChick <- interaction(WS2$fNest, WS2$fChick, drop = TRUE)
MM2_lme <- lme(Beak ~ Age, random = list(fNest = ~1, fNest_fChick = ~1),data = WS2)
summary(MM2_lme)
```

Model validation

```{r}
fit2 <- fitted(MM2_lme)
res2 <- residuals(MM2_lme)

par(mfrow = c(2, 2))
plot(fit2, res2, main = "Residuals vs Fitted")
abline(h = 0, col = "red")
qqnorm(res2)
qqline(res2, col = "red")
hist(res2, breaks = 30, main = "Histogram of Residuals")
```

Compare models: lower AIC indicates better model fit

```{r}
AIC(MM1_lmer, MM2_lmer)
AIC(MM1_lme, MM2_lme)
```

the second model (wuth nested factors) is better for both functions

Anova

```{r}
anova(MM1_lme, MM2_lme)
```

Model MM2_lme is signficantly better

Model validation Left Column (MM1_lme), Right Column (MM2_lme)

```{r}
par(mfcol = c(3, 2), mar = c(4, 4, 2, 1))  

plot(fit, res, main = "Residuals vs Fitted (MM1_lme)") # MM1_lme
abline(h = 0, col = "red")
qqnorm(res, main = "QQ Plot (MM1_lme)")
qqline(res, col = "red")
hist(res, breaks = 30, main = "Histogram (MM1_lme)", col = "gray")

plot(fit2, res2, main = "Residuals vs Fitted (MM2_lme)") # MM2_lme
abline(h = 0, col = "red")
qqnorm(res2, main = "QQ Plot (MM2_lme)")
qqline(res2, col = "red")
hist(res2, breaks = 30, main = "Histogram (MM2_lme)", col = "gray")
```
