---
title: "Exercise_GLM"
author: "Hanh & Verena"
format: html
editor: visual
toc: true
toc-depth: 4
number-sections: true
editor_options: 
  chunk_output_type: console
---

# Exercise for lecture 3 - GLM {.unnumbered}

The main reference, where we also got the data from are:

-   Zuur et al. (2009). Mixed Effects Models and Extensions in Ecology with R\
-   <https:://www.highstat.com>

# Load packages

```{r}
library(dplyr) 
library(lattice)
library(MASS)
```

# Exercise 1 - Binomial GLM for presence - absence data

Load data

```{r}
Tb_data <- read.table("Boar.txt", header = TRUE, sep = "\t")
```

Data source within Zuur et al. (2009): Vicente et al. (2006)

Data check

```{r}
glimpse(Tb_data)
summary(Tb_data)
```

Remove NA in response and predictor variables

```{r}
Tb_data_clean <- Tb_data %>% 
  filter(!is.na(Tb) & !is.na(LengthCT))
```

## Plot relatinoship between Tb ratio vs organism length (LengthCT)

```{r}
par(mfrow = c(1, 2)) # plot two sub plots in one 

boxplot(Tb_data_clean$LengthCT ~ Tb_data_clean$Tb, 
        xlab = "Tuberculosis",
        ylab = "Length",
        main = "(a) Boxplot version",
        border = "black")

plot(Tb_data_clean$LengthCT, Tb_data_clean$Tb, 
     xlab = "Length", 
     ylab = "Tuberculosis", 
     main = "(b) Scatterplot version",
     pch = 1, col = "black")

abline(lm(Tb ~ LengthCT, data = Tb_data_clean), col = "red", lwd = 2) # add linear line
par(mfrow = c(1, 1))
```

## Linear regression

```{r}
lm_model <- lm(Tb ~ LengthCT, data = Tb_data_clean)

summary(lm_model)
```

Model validation plots, with four subplots

```{r}
par(mfrow = c(2, 2))
plot(lm_model) 
par(mfrow = c(1, 1))
```

## Generalized linear regression

```{r}
glm_model <- glm(Tb ~ LengthCT, family = binomial, data = Tb_data_clean)

summary(glm_model)
```

Model validation plots, with four subplots

```{r}
par(mfrow = c(2, 2))
plot(glm_model)
par(mfrow = c(1, 1))
```

ToDo:\
\* Compare the validation plits between the LM and the GLM?\

### Plot: GLM predictions

```{r}
New_data <- data.frame(LengthCT = seq(from = 46.5, to = 165, by = 1)) # create subset of data
Predicted_data <- predict(glm_model, newdata = New_data, type = "response") # get predicted data from model

plot(Tb_data_clean$LengthCT, Tb_data_clean$Tb, 
     xlab = "Length", 
     ylab = "Probability of Tb",
     main = "GLM Fit",
     pch = 1, col = "black")
lines(New_data$LengthCT, Predicted_data, lwd = 2, col = "blue") # add model fit line (non-linear)
```

# Exercise 2 - Poisson GLM for count data

Load data

```{r}
RK_data <- read.table("RoadKills.txt", header = TRUE, sep = "\t")
```

Data check

```{r}
glimpse(RK_data)
summary(RK_data)
names(RK_data)
```

Sampling locations of amphibian kills along the road

```{r}
xyplot(Y ~ X, aspect = "iso", col = 1, pch = 16, data = RK_data)
```

## Plot number of amphibian kills (TOT.N) versus distance to the park (D.PARK)

```{r}
plot(RK_data$D.PARK ~ RK_data$TOT.N,
     xlab="Distance to park (m)",
     ylab="Road kills")
```

## Linear regression

```{r}
Model_1 <- lm(TOT.N ~ D.PARK,data=RK_data)
summary(Model_1) 
```

Model validation plots, with four subplots

```{r}
par(mfrow = c(2, 2))
plot(Model_1)
par(mfrow = c(1, 1))
```

ToDo:\
\* Which model violations do you see?\

## Generalized linear model

```{r}
Model_2<-glm(TOT.N ~ D.PARK,family=poisson,data=RK_data)
summary(Model_2)
```

Model validation plots, with four subplots

```{r}
par(mfrow = c(2, 2))
plot(Model_2)
par(mfrow = c(1, 1))
```

Model evaluation, check for dispersal

```{r}
dispersion_poisson <- sum(residuals(Model_2, type = "pearson")^2) / df.residual(Model_2)
dispersion_poisson
```

Dispersion = 1 (fine)

If Dispersion \> 1.5 (Poisson assumption violated - Overdispersion) --\> Model not complex enough. large parts of variability not explained!

Model predictions

```{r}
MyData <- data.frame(D.PARK = seq(from = 0, to = 25000, by = 1000))
pred <- predict(Model_2, newdata = MyData, type = "link", se.fit = TRUE)
```

Predicted values converted on the scale of response variable

```{r}
fit <- exp(pred$fit)
```

Add upper and lower confidence intervals

```{r}
upper <- exp(pred$fit + 1.96 * pred$se.fit)
lower <- exp(pred$fit - 1.96 * pred$se.fit)
```

### Plot GLM with 95% confidence interval

```{r}
plot(RK_data$D.PARK, RK_data$TOT.N,
     xlab = "Distance to Park (m)",
     ylab = "Road Kills",
     pch = 1, col = "black")
lines(MyData$D.PARK, fit, lty = 1, lwd = 2, col = "darkblue")        # Predicted mean
lines(MyData$D.PARK, upper, lty = 2, col = "blue")                   # Upper CI
lines(MyData$D.PARK, lower, lty = 2, col = "blue")                    # Lower CI
```

# Exercise 3 - Negative binomial GLM for count data

More realistic model: Add explanatory variables

transform data to sqrt

```{r}
RK_data$SQ.POLIC<-sqrt(RK_data$POLIC)
RK_data$SQ.WATRES<-sqrt(RK_data$WAT.RES)
RK_data$SQ.URBAN<-sqrt(RK_data$URBAN)
RK_data$SQ.OLIVE<-sqrt(RK_data$OLIVE)
RK_data$SQ.LPROAD<-sqrt(RK_data$L.P.ROAD)
RK_data$SQ.SHRUB<-sqrt(RK_data$SHRUB)
RK_data$SQ.DWATCOUR<-sqrt(RK_data$D.WAT.COUR)
```

create GLM model with poisson family

```{r}
Model_3 <-glm(TOT.N~D.PARK+OPEN.L+MONT.S+SQ.POLIC+
          SQ.SHRUB+SQ.WATRES+L.WAT.C+SQ.LPROAD+
          SQ.DWATCOUR,family=poisson,data=RK_data)
```

check model output, compare Model_2 and Model_3

```{r}
drop1(Model_3,test="Chi") # drop explanatory variable
AIC(Model_2, Model_3)     # Lower Akaike Information Criterion (AIC) of Model_3 indicates better fit model
summary(Model_3)          # Potential issue: "Dispersion parameter"
```

Check Dispersion

```{r}
dispersion_poisson <- sum(residuals(Model_3, type = "pearson")^2) / df.residual(Model_3)
dispersion_poisson
```

Model still shows overdispersal solution for overdispersed poisson distributed data? -\> use negative binominal distribution

## Negative binomial model

```{r}
Model_4 <-glm.nb(TOT.N ~ D.PARK+OPEN.L+MONT.S+SQ.POLIC+
             SQ.SHRUB+SQ.WATRES+L.WAT.C+SQ.LPROAD+
             SQ.DWATCOUR,link="log",data=RK_data)
summary(Model_4)
```

Check Dispersion

```{r}
dispersion_poisson <- sum(residuals(Model_4, type = "pearson")^2) / df.residual(Model_4)
dispersion_poisson
```

no overdispersal

compare Model_2, Model_3 and Model_4

```{r}
AIC(Model_2, Model_3, Model_4)
```

Lower Akaike Information Criterion (AIC) of Model_4 indicates best fit model

check model performance

```{r}
par(mfrow=c(2,2))
plot(Model_4)
par(mfrow = c(1, 1))
```

# Exercise 4 - abundance in a mesocosm study

some data on a mesocosm study, where we studied drought

read in data

```{r}
drought <- read.table("data_drought.csv", 
                      header = TRUE, 
                      sep = ";", 
                      na.strings = c("NR", "NA"))
```

Investigate effects of different drought scenarios (treatment) on abundance of macroinvertebrates

histogram of reponse variable

```{r}
## Abundance
hist(drought$abundance)

```

fit a linear model

```{r}
lm_model_drought <- lm(abundance ~ treatment, data = drought)
par(mfrow = c(2, 2))
plot(lm_model_drought)
par(mfrow = c(1, 1))
```

ToDo:\
\* Which model violations do you recognise?\
\* What would be your approach to solve this problem?\
\* Please fit a suitable model and check the model fit again. 
