---
title: "In Class Exercise"
author: "Paul Ozoh and Helena Bayat"
format: html
editor: visual
---

# Exercise

You are interested in the following substances with CAS Numbers: "1861-40-1", "161326-34-7", "140923-17-7", "78587-05-0", "13684-56-5". Create a dataframe with the following requirements:

## A. From the EU Pesticide database

From the EU Pesticide database file, get the columns

1.  “Substance”

2.  “Status under Reg. (EC) No 1107/2009”

3.  “RMS”

```{r}

```

### Questions

What are the names of the substances?

What is their status? Are they approved?

What is the RMS (Rapporteur Member State) for each substance?

## B. From the EFSA Open FoodTox database

1.  Filter

“Daphnia magna” from column “Species”

“EC50” from column “Endpoint”

Include columns for “value” and “unit” along with "Species" and "Endpoint"

2.  Perform a left join with the result from Part A. of this exercise

3.  View the result

```{r}

```

5.  Filter

“Daphnia magna” from column “Species”

“NOEC” from column “Endpoint”

Include columns for “value” and “unit” along with "Species" and "Endpoint"

2.  Perform a left join with the result from Part A. of this exercise

3.  View the result

```{r}

```

### Questions

Which substance is most toxic to Daphnia magna?

Does the toxicity seem to relate to the approval of the substance or not?

How does the NOEC compare to the EC50?
