---
title: "In Class Exercise"
author: "Paul Ozoh and Helena Bayat"
format: html
editor: visual
---

# Exercise

You are interested in the following substances with CAS Numbers: "1861-40-1", "161326-34-7", "140923-17-7", "78587-05-0", "13684-56-5". Create a dataframe with the following requirements:

## A. From the EU Pesticide database

From the EU Pesticide database file, get the columns

1.  “Substance”

2.  “Status under Reg. (EC) No 1107/2009”

3.  “RMS”

```{r}
# CAS numbers to look up
required_cas <- c("1861-40-1", "161326-34-7", "140923-17-7", "78587-05-0", "13684-56-5")

# Filter the dataset and select the columns of interest
filtered_data <- df %>%
  filter(`CAS Number` %in% required_cas) %>%
  select(`CAS Number`, Substance, `Status under Reg. (EC) No 1107/2009`, RMS)

# View the filtered data
print(filtered_data)

```

### Questions

What are the names of the substances?

What is their status? Are they approved?

What is the RMS (Rapporteur Member State) for each substance?

## B. From the EFSA Open FoodTox database

1.  Filter

“Daphnia magna” from column “Species”

“EC50” from column “Endpoint”

Include columns for “value” and “unit” along with "Species" and "Endpoint"

2.  Perform a left join with the result from Part A. of this exercise

3.  View the result

```{r}
# Filter the toxicity data for Daphnia magna + NOEC
tox_data_added <- data %>%
  filter(Species == "Daphnia magna", Endpoint == "EC50") %>%
  select(Substance, Species, Endpoint, value, unit)

# Step 2: Join with filtered pesticide data
combined_data <- filtered_data %>%
  left_join(tox_data_added, by = "Substance")

# Step 3: View result
print(combined_data)

```

5.  Filter

“Daphnia magna” from column “Species”

“NOEC” from column “Endpoint”

Include columns for “value” and “unit” along with "Species" and "Endpoint"

2.  Perform a left join with the result from Part A. of this exercise

3.  View the result

```{r}
# Filter the toxicity data for Daphnia magna + NOEC
tox_data_added <- data %>%
  filter(Species == "Daphnia magna", Endpoint == "NOEC") %>%
  select(Substance, Species, Endpoint, value, unit)

# Step 2: Join with filtered pesticide data
combined_data <- filtered_data %>%
  left_join(tox_data_added, by = "Substance")

# Step 3: View result
print(combined_data)

```

### Questions

Which substance is most toxic to Daphnia magna?

Does the toxicity seem to relate to the approval of the substance or not?

How does the NOEC compare to the EC50?
