---
title: "Ecotoxicological Databases Exercise"
author: "Helena S. Bayat"
format: html
editor: visual
---

## Load packages

```{r}

#| label: setup
library(standartox)
library(dplyr)
library(ssdtools)
library(ggplot2)

```

## Run a Standartox query

Let's start with copper sulfate, a widely used fungicide. We use the CAS number to filter for this chemical. We also filter by endpoint and effect, to select only LC50s or EC50s with mortality. We select a duration of 24 hours, commonly used in acute toxicity testing.

```{r}

# query
copper = stx_query(cas = '7758-98-7',
              endpoint_group = 'XX50',
                effect = 'mortality',
              concentration_unit = 'g/l',
              duration = 24)

```

## Explore the data

Lets take a look at our data. We use the group_by() and count() functions of the dplyr package to count how many data points we have per species.

```{r}

# Data points per species
species_c <- copper %>% group_by(tax_taxon) %>% count()
species_c

```

Now, using the same functions, let's check how many data points we have per group.

```{r}

# Data points per group
group_c <- copper %>% group_by(tax_group) %>% count()
group_c

```

To count the number of species per group, we repeat the functions using the pipe (%\>%).

```{r}

# Number of species per group
num_c <- copper %>% group_by(tax_taxon, tax_group) %>% count() %>% group_by(tax_group) %>% count()
num_c

```

## Species sensitivity distribution and HC5

Use the Standartox data to plot a species sensitivity distribution and calculate HC5 for each chemical.

First, we change the column names to be compatible with the ssdtools package.

```{r}

# change column names
colnames(copper)[7] <- "Conc"
colnames(copper)[23] <- "Species"

```

Then we trim the data to one value per species - in this case we select the median value.

```{r}

# select median value for each species
copper <- copper %>% group_by(Species) %>% slice_min(abs(Conc - median(Conc)), n = 1, with_ties = FALSE)

```

Now we fit the distribution to the data.

```{r}

# fit distribution
ssd_c <- ssd_fit_dists(copper)

```

Then we plot our result.

```{r}

# plot
ssd_plot_cdf(ssd_c)

```

Finally, we calculate a HC5 from our fitted distribution.

HC5 is the concentration at which 5% of the species are affected; the HC stands for "hazard concentration". This value can be calculated using the fitted species sensitivity distribution. The distribution can also be used to calculate other hazard concentrations, such as HC10, HC20, or HC50.

```{r}

# HC5
hc5_c <- ssd_hc(ssd_c, proportion = 0.05, ci = TRUE)
hc5_c$est

```

## Repeat steps for DDT, glyphosate, and imidacloprid

Run a Standartox query, look at the data, plot a SSD, and calculate HC5 for each chemical.

### DDT

Standartox query

```{r}

#query
ddt = stx_query(cas = '50-29-3',
              endpoint_group = 'XX50',
                effect = 'mortality',
              concentration_unit = 'g/l',
              duration = 24)

```

Explore the data

```{r}

# Data points per species
species_d <- ddt %>% group_by(tax_taxon) %>% count()

# Data points per group
group_d <- ddt %>% group_by(tax_group) %>% count()

# Number of species per group
num_d <- ddt %>% group_by(tax_taxon, tax_group) %>% count() %>% group_by(tax_group) %>% count()

```

Plot SSD

```{r}

# change column names
colnames(ddt)[7] <- "Conc"
colnames(ddt)[23] <- "Species"

# select median value for each species
ddt <- ddt %>% group_by(Species) %>% slice_min(abs(Conc - median(Conc)), n = 1, with_ties = FALSE)

# fit distribution
ssd_ddt <- ssd_fit_dists(ddt)
  
# plot
ssd_plot_cdf(ssd_ddt)

```

Calculate HC5

```{r}

# HC5
hc5_ddt <- ssd_hc(ssd_ddt, proportion = 0.05, ci = TRUE)
hc5_ddt$est

```

### Glyphosate

Standartox query

```{r}

#query
glyphosate = stx_query(cas = '1071-83-6',
              endpoint_group = 'XX50',
                effect = 'mortality',
              concentration_unit = 'g/l',
              duration = 24)

```

Explore the data

```{r}

# Data points per species
species_g <- glyphosate %>% group_by(tax_taxon) %>% count()

# Data points per group
group_g <- glyphosate %>% group_by(tax_group) %>% count()

# Number of species per group
num_g <- glyphosate %>% group_by(tax_taxon, tax_group) %>% count() %>% group_by(tax_group) %>% count()

```

Plot SSD

```{r}

# change column names
colnames(glyphosate)[7] <- "Conc"
colnames(glyphosate)[23] <- "Species"

# select median value for each species
glyphosate <- glyphosate %>% group_by(Species) %>% slice_min(abs(Conc - median(Conc)), n = 1, with_ties = FALSE)

# fit distribution
ssd_g <- ssd_fit_dists(glyphosate)
  
# plot
ssd_plot_cdf(ssd_g)

```

Calculate HC5

```{r}

# HC5
hc5_g <- ssd_hc(ssd_g, proportion = 0.05, ci = TRUE)
hc5_g$est

```

### Imidacloprid

Standartox query

```{r}

#query
imidacloprid = stx_query(cas = '138261-41-3',
              endpoint_group = 'XX50',
              effect = 'mortality',
              concentration_unit = 'g/l',
              duration = 24)

```

Explore the data

```{r}

# Data points per species
species_i <- imidacloprid %>% group_by(tax_taxon) %>% count()

# Data points per group
group_i <- imidacloprid %>% group_by(tax_group) %>% count()

# Number of species per group
num_i <- imidacloprid %>% group_by(tax_taxon, tax_group) %>% count() %>% group_by(tax_group) %>% count()

```

Plot SSD

```{r}

# change column names
colnames(imidacloprid)[7] <- "Conc"
colnames(imidacloprid)[23] <- "Species"

# select median value for each species
imidacloprid <- imidacloprid %>% group_by(Species) %>% slice_min(abs(Conc - median(Conc)), n = 1, with_ties = FALSE)

# fit distribution
ssd_im <- ssd_fit_dists(imidacloprid)
  
# plot
ssd_plot_cdf(ssd_im)

```

Calculate HC5

```{r}

# HC5
hc5_im <- ssd_hc(ssd_im, proportion = 0.05, ci = TRUE)
hc5_im$est

```

## Save files

We save the files so that we can look at the results and continue working with the data without rerunning a Standartox query. For a data analysis with many steps, this saves a lot of time and computational power.

The files will be saved to the current working directory of R. The default working directory is the folder in which this exercise file is saved.

```{r}

# write files

write.csv(copper, "copper.csv", row.names = FALSE, na = "")
write.csv(ddt, "ddt.csv", row.names = FALSE, na = "")
write.csv(glyphosate, "glyphosate.csv", row.names = FALSE, na = "")
write.csv(imidacloprid, "imidacloprid.csv", row.names = FALSE, na = "")

```

## Questions

Which chemical has the lowest HC5? Which chemical has the highest?

Which chemical is most toxic?

which chemical has the most data for fish?

Which chemical has the most data for invertebrates?
