------------------------------------------------------------------------

------------------------------------------------------------------------

---
title: "Lambda-cyhalothrin demonstration: aquatic species sensitivity and risk"
author: "Paul Ozoh"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This short demonstration shows how ecotoxicological database records can be filtered and used to:

1.  Rank aquatic species by chronic sensitivity, and
2.  Calculate organism-level risk quotients (RQ) using a simplified Tier 1 approach.

The demonstration uses **lambda-cyhalothrin** and a **Step 1 PECsw of 0.23 µg/L**.

## Key assumptions

-   Data source: `ReferencePoints_KJ_2023.xlsx`
-   Endpoint restriction: **NOEC** only
-   Units are harmonised to **µg/L**
-   Exposure route was reported as **"Not reported"**, so aquatic records are identified by species-based classification.
-   Algae are excluded from this demonstration because the dataset was filtered to NOEC only, whereas algal data are often reported as growth-related endpoints such as ErC50/EbC50 or EC10.
-   Simplified Tier 1 RAC derivation for each organism:

$$
RAC = \frac{NOEC}{10}
$$

-   Risk quotient:

$$
RQ = \frac{PEC_{sw}}{RAC}
$$

## Load packages and data

```{r}
library(readxl)
library(dplyr)
library(ggplot2)

pecsw <- 0.23

raw_data <- read_excel("ReferencePoints_KJ_2023.xlsx")

names(raw_data) <- tolower(names(raw_data))

data_clean <- raw_data
```

## Filter lambda-cyhalothrin NOEC records

```{r}
lambda_chronic <- data_clean %>%
  filter(grepl("lambda", substance, ignore.case = TRUE)) %>%
  filter(endpoint == "NOEC") %>%
  mutate(
    value_ugl = case_when(
      grepl("mg", unit, ignore.case = TRUE) ~ value * 1000,
      grepl("µg|ug", unit, ignore.case = TRUE) ~ value,
      TRUE ~ NA_real_
    )
  )

lambda_chronic %>%
  select(substance, species, endpoint, value, unit, value_ugl)
```

## Keep aquatic organisms only.

Because `NOEC` is reported for aquatic and terrestrial organisms, aquatic records are selected manually from the species list.

```{r}
unique(lambda_chronic$species)
```

```{r}
lambda_aquatic <- lambda_chronic %>%
  filter(species %in% c("Fathead minnow", "Harlequin fly", "Mysid shrimp")) %>%
  mutate(
    group = case_when(
      species == "Fathead minnow" ~ "fish",
      species %in% c("Harlequin fly", "Mysid shrimp") ~ "invertebrates",
      TRUE ~ "other"
    )
  )

lambda_aquatic %>%
  select(species, group, value_ugl)
```

## Sensitivity ranking

Lower NOEC values indicate higher sensitivity.

```{r}
lambda_aquatic %>%
  arrange(value_ugl) %>%
  mutate(rank = row_number()) %>%
  select(rank, species, group, value_ugl)
```

```{r}
#| label: sensitivity-plot
#| fig-width: 9
#| fig-height: 5

lambda_aquatic <- lambda_aquatic %>%
  rename(value_ugL = value_ugl)

ggplot(lambda_aquatic,
       aes(x = reorder(species, value_ugL), y = value_ugL, color = group)) +
  geom_point(size = 4) +
  coord_flip() +
  scale_y_log10() +
  labs(
    title = "Aquatic Species Sensitivity Ranking (Lambda-cyhalothrin)",
    x = "Species",
    y = "NOEC (µg/L, log scale)"
  ) +
  theme_minimal()
```

### Interpretation

The most sensitive aquatic species is expected to drive the regulatory conclusion in a simplified Tier 1 screen. Here, **Mysid shrimp** is the critical species because it has the lowest NOEC.

## Calculate RAC and RQ for each organism

```{r}

lambda_rq <- lambda_aquatic %>%
  mutate(
    rac = value_ugL / 10,
    rq = pecsw / rac,
    risk_status = if_else(rq < 1, "Acceptable", "Unacceptable")
  ) %>%
  arrange(desc(rq))

lambda_rq %>%
  select(species, group, value_ugL, rac, rq, risk_status) %>%
  mutate(
    value_ugL = signif(value_ugL, 3),
    rac = signif(rac, 3),
    rq = signif(rq, 4)
  )
```

```{r}

#| fig-width: 9
#| fig-height: 5

p_rq <- ggplot(lambda_rq, aes(x = reorder(species, rq), y = rq, fill = risk_status)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 1) +
  geom_text(aes(label = round(rq, 1)), hjust = -0.1, size = 4) +
  coord_flip(clip = "off") +
  scale_y_log10(expand = expansion(mult = c(0.05, 0.25))) +
  labs(
    title = "Risk quotient by aquatic organism (lambda-cyhalothrin)",
    x = NULL,
    y = "RQ = PECsw / RAC (log scale)",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = margin(10, 60, 10, 10),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

p_rq
```

## Summary of results

```{r}
summary_table <- lambda_rq %>%
  transmute(
    species,
    group,
    noec_ugL = signif(value_ugL, 3),
    rac_ugL = signif(rac, 3),
    pecsw_ugL = pecsw,
    rq = signif(rq, 4),
    risk_status
  )

summary_table
```

## Take-home message

-   The aquatic sensitivity ranking shows that **Mysid shrimp** is the critical species.
-   The simplified Tier 1 RAC values lead to **RQ \> 1 for all three organisms**.
-   In this demonstration, **no organism shows acceptable risk at Step 1**.
-   The comparison also shows how species sensitivity translates directly into risk characterisation.

## 
