---
title: "IBM-Daphnia-model"
author: "Based on Petzoldt & Rinke (2007)"
date: "2026-07-08"
output: 
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document presents an Individual-Based Model (IBM) for Daphnia population dynamics using the simecol package in R. The model simulates growth and egg production, mortality, and hatching processes in Daphnia populations.

**Reference:** Petzoldt & Rinke (2007) simecol: An Object-Oriented Framework for Ecological Modeling in R. Journal of Statistical Software. <https://www.jstatsoft.org/v22/i09>.

## Load Libraries

```{r libraries}
library(.............)
```

## IBM-Daphnia Model Setup

The IBM-Daphnia model is implemented as a simecol object with three main processes: growth and egg production, mortality, and hatching of neonates. 

```{r model-setup}
# A simecol object
ibm_daphnia <- new(..................,   # Built-in class
  main = function(time, init, parms) {
    # Three main processes per time step
    init <- .......(init, parms)         # Growth and egg production
    init <- .......(init, parms)         # Mortality 
    init <- .......(init, parms)         # Hatching of new neonates
    init                              
  },

  # Model parameters (constants)
  parms = list(
    # Growth parameters 
    a1 = 1.167,        # mm, max size gain from food
    a2 = 0.573,        # mg/L, half-saturation food level
    a3 = 1.420,        # mm, baseline asymptotic size offset
    a4 = 2.397,        # d, term linking growth rate to max individual size
    b1 = 1.089e-2,     # d^-1, growth-rate coefficient
    b2 = 0.122,        # 1/°C, growth rate sensitivity to temperature

    # Clutch size parameters 
    X_max_slope = 23.83,     # eggs, max slope of eggs
    K_s_slope   = 0.65,      # mg/L, half-saturation food level
    beta_min    = -29.28,    # eggs, negative intercept 
    u_c         = 1,         # L/mg, food effect on clutch

    # IBM model parameters
    L_0_Hall = 0.35,   # mm, size of neonates (reference of Hall data) 
    L_0      = ....,   # mm, size of neonates at birth (in this model)
    SAM      = ....,   # mm, size at maturity (SAM, below this threshold: no reproduction)
    maxage   = ....,     # d, baseline death age

    # Environmental conditions
    temp = ......,         # °C, water temperature
    food = .......         # mg/L, food (algae) concentration 
  ),

  # The starting population
  init = data.frame(age = .....,       # d, age of a newborn neonate
                    size = ....,       # mm, size of a newborn neonate  
                    eggs = ....,       # eggs carried in brood pouch
                    eggage = ..),      # d, age of the current egg clutch 
  
  # Simulation time (from day 0 to day 60, time step = 1 day)
  times = c(from = ....., to = ....., by = .....),   
  
  # Solver 
  solver = "myiteration",          # own customer solver (defined in the next chunk)
  
  # Equations for live, survive, and hatching processes 
  equations = list()               # empty list (filled in the next chunk)
)
```

## Model Equations

The `equations` slot holds functions, that are called by the model in each time step: the growth and egg reproduction (`tefi`), the egg-development (`bottrell`), the neonate (`newdaphnia`), and the three processes (`live`, `survive`, `hatch`).

```{r equations}
equations(.............) <- list(
  # New neonates: build new offspring rows
  newdaphnia = function(SON, n) {       # create n new neonates, each of size SON (size of neonate)
    if (n > 0) {                        # If condition: only if at least 1 egg hatched
      data.frame(age = rep(0, n),       # newborn at age = 0
                 size = SON,            # born at size L_0
                 eggs = 0,              # carry no egg yet
                 eggage = 0)            # no clutch yet
    } else {
      NULL
      }
  },

  # Egg development time as function of temperature 
  bottrell = function(temp) {           # Bontrell function: number of days eggs needs to develop before hatching
    exp(3.3956 + 0.2193 * log(temp) - 0.3414 * log(temp)^2)
  },

  # Temperature-Food Interaction model (TeFI)
  tefi = function(time, temp, food, parms){
    with(parms, {
      deltaL <- L_0 - L_0_Hall
      k <- b1 * exp(b2 * temp)                                      # Growth rate k rises with temperature
      L_max <- (a1 * food)/(a2 + food) + a3 - k * a4                # Max body size L_max. Food enters as a saturating Michaelis-Mentel term.          
      L <- L_max - (L_max - L_0_Hall) * exp (-k * time) + deltaL    # Length L approaches L_max over time
      E <- (X_max_slope * food)/(K_s_slope + food) * L + beta_min * (1 - exp(-u_c * food))  # Clutch size E as function of food and body length

      as.data.frame(cbind(L, E))
    })
  },

  #1 - Individual growth and egg production
  live = function(inds, parms){
    with(parms,{
      inds$age <- inds$age + DELTAT
      inds$eggage <- ifelse(inds$size > ...... & inds$eggs > ........, inds$eggage + DELTAT, 0)

      tefi_out <- tefi(inds$age, temp, food, parms)
      inds$size <- tefi_out$L
      neweggs <- round(tefi_out$E)

      inds$eggs <- ifelse(inds$size > ......... & inds$eggage == ......., neweggs, inds$eggs)
      ........
    })
  },

  #2- Mortality 
  survive = function(inds, parms) subset(inds, inds$age < ..........),

  #3- Hatching of new neonates 
  hatch = function(inds, parms) {
    with(parms, {
      edt <- bottrell(temp)                                      # egg development time
      have.neo <- which(inds$eggs >..... & inds$eggage > .....)  # conditions: have eggs that finished developing
      eggs <- inds$eggs[have.neo]                                # eggs carried by those satisfying above conditions
      new.neo <- sum(eggs)                                       # total neonates born

      inds$eggs[have.neo] <- inds$eggage[have.neo] <- 0      # clutches released, egg and eggage reset to 0
      newinds <- newdaphnia(L_0, new.neo)                    # newborns
      rbind(inds, .......)                                   # newborns are added to the population
    })
  }
)
```

## Custom Solver

The model uses a custom iterator solver for the IBM-Daphnia model.

```{r solver}
myiteration <- function(y, ...) {
  observer <- function(res) {
    c(number = nrow(res),             # number of individuals in the population
      meansize = mean(res$.......),   # mean size of individuals in the population
      meanage = mean(res$........),   # mean age of individuals in the population
      meaneggs = mean(res$.......))   # mean eggs per individual
  }

  init <- y@init
  times <- fromtoby(y@times)
  func <- y@main
  parms <- y@parms
  equations <- y@equations
  equations <- addtoenv(equations)
  environment(func) <- environment()
  parms$DELTAT <- 0
  res <- observer(init)
  out <- res

  for (i in 2:length(times)) {
    time <- times[i]
    parms$DELTAT <- times[i] - times[i - 1]
    init <- func(time, init, parms)
    res <- observer(init)
    out <- rbind(out, res)
  }

  row.names(out) <- NULL
  out <- cbind(times, out)
  as.data.frame(out)
}
```

## Model Simulation

Run the IBM-Daphnia model simulation using sim(simObj) and examine the results.

```{r simulation}
# Model simulation
ibm_daphnia <- sim(.........)

# Model results
head(ibm_daphnia@out)

```

## Population Abundance Patterns

Visualize the Daphnia population abundance patterns over time.

```{r abundance-plots, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))
plot(ibm_daphnia@out[,1],ibm_daphnia@out[,2], 
     xlab="Time", ylab="Abundance",  
     type="l", col = "darkgreen", lty=2, lwd=2)
plot(ibm_daphnia@out[,1],ibm_daphnia@out[,2], 
     xlab="Time", ylab="log (Abundance)", log = "y", 
     type="l", col = "blue", lty=2, lwd=2)
```

## Custom Plot Method

Create a custom plot method for the individual-based model that shows age, eggs, and abundance patterns.

```{r custom-plot-method}
setMethod("plot", c("indbasedModel", "missing"), function(x, y, ...) {
  o <- out(x)
  par(mfrow = c(2, 2))
  plot(o$times, o$meanage, type = "l", xlab = "Time", ylab = "Mean age (d)", ylim=c(0,12), col = "grey", lty=1, lwd=2)
  plot(o$times, o$meaneggs, type = "l", xlab = "Time", ylab = "Eggs per individual", ylim=c(0,5), col = "orange", lty=1, lwd=2)
  plot(o$times, o$number, type = "l", xlab = "Time", ylab = "Abundance", col = "darkgreen", lty=1, lwd=2)
  plot(o$times, o$number, type = "l", xlab = "Time", ylab = "Abundance (log scale)", log = "y", col = "blue", lty=1, lwd=2)
})
```

```{r custom-plots, fig.width=10, fig.height=8}
plot(ibm_daphnia)
```

## Mortality Scenarios: Size-Selective Predation

### Scenario Setup

We examine four different mortality scenarios:

-   **Sc0**: No mortality (baseline case with maximum growth potential)
-   **Sc1**: Random mortality (constant per capita mortality, independent of prey size)
-   **Sc2**: Size-selective mortality (small individuals preferred - invertebrate-like predation)
-   **Sc3**: Size-selective mortality (large individuals preferred - fish-like predation)

**Main hypothesis:** Predators preferring large Daphnia preys increases Daphnia population mortality more than non-selective (random) predation.

```{r scenario-setup}
# Setup for all three scenarios
...... <- ibm_daphnia
times(Sc0) <- c(from = 0, to = 30, by = 0.2)
parms(Sc0)[c("temp", "food", "mort")] <- c(15, 0.4, 0.1)  # Added 'mort' to parms
init(Sc0) <- data.frame(age = rep(10, 50), size = rep(2.5, 50),
                        eggs = rep(5, 50), eggage = runif(50, 0, 4))

# Replace default survive function with user-defined size-dependent mortality function fmort 
equations(Sc0)$survive <- function(inds, parms) {
  abundance <- nrow(inds)
  rnd <- runif(abundance)
  mort <- fmort(parms$mort, inds$size) * parms$DELTAT
  subset(inds, rnd > mort)
}

# Define mortality (fmort functions) for three scenarios (with shared Sc0 setup, but differ in fmort())
Sc1 <- Sc2 <- Sc3 <- .......

# Scenario-specific mortality functions
# No mortality
equations(Sc0)$fmort <- function(mort, x) .....
# Constant mortality
equations(Sc1)$fmort <- function(mort, x) .....
# Small individuals die more
equations(Sc2)$fmort <- function(mort, x) ..... * 2 * rank(-x)/(length(x)+1)  
# Large individuals die more
equations(Sc3)$fmort <- function(mort, x) ..... * 2 * rank(x)/(length(x)+1) 
```

### Scenario Simulations

```{r scenario-simulations}
# Run simulations for scenarios
scenarios <- list(Sc0 = Sc0, Sc1 = Sc1, Sc2 = Sc2, Sc3 = Sc3)
results <- lapply(scenarios, sim)
```

### Scenario Results Visualization

```{r scenario-plots, fig.width=10, fig.height=6}
# Plot scenario results
par(mfrow = c(1,1))

colors <- c("black", "blue", "orange", "red")
labels <- c("Sc0: No mortality", "Sc1: Random mortality", 
            "Sc2: Small prey preferred", "Sc3: Large prey preferred")

plot(NULL, xlim = c(0, 30), ylim = c(10, 5000), log = "y",
     xlab = "Time (days)", ylab = "Abundance (log scale)", main = "Daphnia IBM: Mortality Scenarios")
for (i in 1:4) {
  lines(out(results[[i]])$times, out(results[[i]])$number, col = colors[i], lwd = 2)
}
legend("topleft", legend = labels, col = colors, lty = 1, lwd = 2)
```

## Questions

### Question 1: Compare results of different scenarios Sc0 to Sc4

-   **Sc0 (No mortality):** [Add your answer here]
-   **Sc1 (Random mortality):** [Add your answer here]
-   **Sc2 (Small prey preferred):** [Add your answer here]
-   **Sc3 (Large prey preferred):** [Add your answer here]

### Question 2: Hypothesis Testing

Does the scenario results support the hypothesis that size-selective predation targeting adults leads to stronger population suppression?

[Add your answer here]
