---
title: "TKTD-IBM Exercise — Copper Effects on a *Daphnia* population"
subtitle: "Coupled TKTD - IBM Daphnia magna model: DEBtox (sublethal) + GUTS (lethal)"
author: "Based on open data from Vlaeminck et al. (2021) *Environ Toxicol Chem* 40:513-527, doi:10.1002/etc.4946"
date: "2026"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true

---
## Introduction

The exercise provides the script to run an example of a coupled **TKTD-IBM** model:
a toxicokinetic-toxicodynamic damage model feeding **DEBtox** (sublethal) and **GUTS** (lethal) 
effects into an IBM. This is different to the *original* Vlaeminck et al. (2021) model, which is
a full DEB-IBM. This exercise includes their exact toxicology values, but simplifies the energy budget.
Implementation for teaching purposes only.

**Reference.** Vlaeminck K., Viaene K.P.J., Van Sprang P., De Schamphelaere K.A.C. (2021). *Development and
Validation of a Mixture Toxicity Implementation in the Dynamic Energy Budget–Individual-Based Model:
Effects of Copper and Zinc on Daphnia magna Populations.* Environ Toxicol Chem 40(2):513–527.
doi:10.1002/etc.4946.

**Data availability.** The `.xlsx` files used in this exercise are subsets of the full
supplementary dataset for Vlaeminck et al. (2021). Full deposited dataset (8 parts, Figshare):
https://doi.org/10.6084/m9.figshare.11799198; https://doi.org/10.6084/m9.figshare.11799342; 
https://doi.org/10.6084/m9.figshare.11799357; https://doi.org/10.6084/m9.figshare.11799612; https://doi.org/10.6084/m9.figshare.11799735; https://doi.org/10.6084/m9.figshare.11799768; https://doi.org/10.6084/m9.figshare.11799804; https://doi.org/10.6084/m9.figshare.11799864. 

## Research questions

1. **Sublethal (DEBtox):** How does copper, by **reducing assimilation**, slow individual growth and reproduction of individuals, 
   and how visible is that at the population level?
2. **Lethal (GUTS):** Above a damage threshold, copper causes **stochastic death** of individuals. How does it affect population trajectory and structure?
3. **Validation:** Do the IBM predictions, driven only by **independently calibrated** copper parameters, reproduce the **observed** population experiment?

---

# Load library

```{r libraries}
library("simecol")   
library("readxl")    
```

---

# DEBtox (sublethal): 

```{r read-debtox}
path_ind <- "SI_individual_experiment.xlsx"   
pal <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")  

deb <- read_excel(path_ind, sheet = "DEBtox_calibration_data", skip = 8)
names(deb)[1:6] <- c("Treatment", "Day", "Me.Conc", "Nom.Conc", "Reproduction", "Size")
deb <- deb[, 1:6]
deb <- deb[!is.na(deb$Treatment) & deb$Treatment != "-", ]
deb$Day          <- as.numeric(deb$Day)
deb$Me.Conc      <- as.numeric(deb$Me.Conc)
deb$Reproduction <- suppressWarnings(as.numeric(deb$Reproduction))
deb$Size         <- suppressWarnings(as.numeric(deb$Size))   

ctrl     <- subset(deb, Treatment == "Control") 
ctrl <- ctrl[order(ctrl$Day), ]
cu_deb   <- subset(deb, Treatment == "Copper")
cu_concs <- sort(unique(cu_deb$Me.Conc))        
size0    <- ctrl$Size[ctrl$Day == 0] / 1000      

par(mfrow = c(1, 2))
plot(ctrl$Day, ctrl$Reproduction, type = "l", lwd = 2, col = "black",
     xlab = "Time (days)", ylab = "Cumulative reproduction (#)",
     main = "Sublethal: reproduction", ylim = c(0, 60))
for (i in seq_along(cu_concs)) {
  d <- subset(cu_deb, Me.Conc == cu_concs[i]); d <- d[order(d$Day), ]
  lines(d$Day, d$Reproduction, col = pal[i], lwd = 2)
}
legend("topleft", legend = c("Control", paste0(cu_concs, " ug/L")),
       col = c("black", pal), lwd = 2, bty = "n", cex = 0.8)

# Body length (measured on days 0, 7, 14, 21 for control; for Cu lengths were only measured from day 7, 14, 21) 
ctrl_g <- ctrl[!is.na(ctrl$Size), ]
plot(ctrl_g$Day, ctrl_g$Size / 1000, type = "b", pch = 19, lwd = 2, col = "black",
     xlab = "Time (days)", ylab = "Body length (mm)", ylim = c(0.5, 4),
     main = "Sublethal: growth")
for (i in seq_along(cu_concs)) {
  d <- subset(cu_deb, Me.Conc == cu_concs[i]); d <- d[!is.na(d$Size), ]; d <- d[order(d$Day), ]
  if (nrow(d) > 0) {                                  
    xx <- c(0, d$Day); 
    yy <- c(size0, d$Size / 1000)  
    lines(xx, yy, type = "b", pch = 19, col = pal[i], lwd = 2)
  }
}
legend("bottomright", legend = c("Control", paste0(cu_concs, " ug/L")),
       col = c("black", pal), lwd = 2, bty = "n", cex = 0.75)
```

---

# GUTS (lethal): 

```{r read-guts}
guts <- read_excel(path_ind, sheet = "GUTS_calibration_data", skip = 6)
names(guts)[1:5] <- c("Treatment", "Day", "Me.Conc", "Nom.Conc", "Survival")
guts <- guts[, 1:5]
guts <- guts[!is.na(guts$Treatment) & guts$Treatment != "-", ]   
guts$Day      <- as.numeric(guts$Day)
guts$Me.Conc  <- as.numeric(guts$Me.Conc)
guts$Survival <- as.numeric(guts$Survival)

cu_surv <- subset(guts, Treatment == "Copper")   
```

```{r guts-function}
# Damage Df builds toward Cw
# Once it passes threshold zw the hazard H switches on.
# Survival is exp(-H).
guts_survival <- function(Cw, t, kD = 0.4057, zw = 38.235, bw = 1.0277e-2) {
  tf <- seq(0, max(t), by = 1/24)                                  # hourly grid for the integral
  Df <- Cw * (1 - exp(-kD * tf))                                   # analytical damage trajectory
  hf <- bw * pmax(0, Df - zw)                                      # instantaneous hazard
  H  <- c(0, cumsum((head(hf, -1) + tail(hf, -1)) / 2 * diff(tf))) # cumulative hazard 
  approx(tf, exp(-H), xout = t)$y                                  # survival at requested days
}
```

```{r guts-plot}
plot(NULL, xlim = c(0, 21), ylim = c(0, 1),
     xlab = "Time (days)", ylab = "Survival fraction",
     main = "Observed survival (points) vs GUTS prediction (lines)")
for (i in seq_along(cu_concs)) {
  d <- subset(cu_surv, Me.Conc == cu_concs[i]); d <- d[order(d$Day), ]
  points(d$Day, d$Survival, col = pal[i], pch = 19)               
  tt <- seq(0, 21, by = 0.5)
  lines(tt, guts_survival(cu_concs[i], tt), col = pal[i], lwd = 2) 
}
legend("bottomleft", legend = paste0(cu_concs, " ug/L"),
       col = pal, pch = 19, lwd = 2, bty = "n", cex = 0.9)
```

---

# Simplified (modified for teaching purpose) TKTD-IBM: 

```{r model-object}
ibm_cu <- new("indbasedModel",
  main = function(time, init, parms) {
    init <- live(init, parms)     # growth + egg production + (added) copper DAMAGE UPTAKE + (added) DEBtox
    init <- survive(init, parms)  # (added) GUTS + (added) starvation + maximum age
    init <- hatch(init, parms)    # neonates born (kept as in Exercise 2)
    init
  },

  parms = list(
    # Growth parameters 
    a1 = 1.167, a2 = 0.573, a3 = 1.420, a4 = 2.397, b1 = 1.089e-2, b2 = 0.122,
    # Clutch-size parameters 
    X_max_slope = 23.83, K_s_slope = 0.65, beta_min = -29.28, u_c = 1,
    # Life-history parameters 
    L_0_Hall = 0.35, L_0 = 0.65, SAM = 1.50, maxage = 60,
    # Environmental conditions
    temp = 20,        # degree C
    food_max = 0.5,   # Max food (mg/L)

    # Density-dependent food ~ starvation proxy (simplification for course exercise)
    Nhalf = 250,  # density at which available food halves ~ defined as the carrying capacity

    # Copper exposure + calibrated TKTD (Copper_gekal_assim.txt / Table 4) 
    Cw = 0,                       # external copper concentration (µg/L); 0 = clean control
    kD = 0.01690427 * 24,         # damage rate
    NEC = 12.35223,               # DEBtox no-effect concentration 
    cT = 0.01143088,              # DEBtox stress slope/multiplier
    zw = 38.23527,                # GUTS damage threshold (m_n)
    bw = 4.28189e-4 * 24,         # GUTS killing rate
    mortconst = 1.013762e-3 * 24  # starvation constant
  ),

  # Initial population: 6 brooding adults + 10 neonates
  init = local({
    set.seed(42)
    adults   <- data.frame(age = 8, size = 2.2, eggs = 3, eggage = runif(6, 0, 3), Dw = 0)
    neonates <- data.frame(age = 0, size = 0.65, eggs = 0, eggage = 0, Dw = 0)[rep(1, 10), ]
    rbind(adults, neonates)
  }),

  times = c(from = 0, to = 75, by = 1),   # 75-day population experiment
  solver = "myiteration",
  equations = list()
)
```

---

# The equations: 

```{r equations}
equations(ibm_cu) <- list(

  # new neonates: a Dw = 0 (born undamaged)
  newdaphnia = function(SON, n) {
    if (n > 0) data.frame(age = rep(0, n), size = SON, eggs = 0, eggage = 0, Dw = 0) else NULL
  },

  # egg development time 
  bottrell = function(temp) exp(3.3956 + 0.2193 * log(temp) - 0.3414 * log(temp)^2),

  # Temperature-Food interaction growth
  # Note: the parameter maximum food is named `food_max`, NOT `food`. 
  tefi = function(time, temp, food, parms) {
    with(parms, {
      deltaL <- L_0 - L_0_Hall
      k <- b1 * exp(b2 * temp)
      L_max <- (a1 * food) / (a2 + food) + a3 - k * a4
      L <- L_max - (L_max - L_0_Hall) * exp(-k * time) + deltaL
      E <- (X_max_slope * food) / (K_s_slope + food) * L + beta_min * (1 - exp(-u_c * food))
      as.data.frame(cbind(L, E))
    })
  },

  #1 LIVE — growth, egg production, copper damage uptake and DEBtox 
  live = function(inds, parms) {
    with(parms, {
      N <- nrow(inds)

      # Damage
      inds$Dw <- inds$Dw + kD * (Cw - inds$Dw) * DELTAT
      inds$Dw <- pmin(inds$Dw, Cw)                       

      # Effective food = density-dependent supply 
      food_dd  <- food_max / (1 + N / Nhalf)             # density-dependent food
      s        <- cT * pmax(0, inds$Dw - NEC)            # DEBtox stress 
      food_eff <- food_dd / (1 + s)                      # assimilation PMoA
      
      # Growth & reproduction 
      inds$age    <- inds$age + DELTAT
      inds$eggage <- ifelse(inds$size > SAM & inds$eggs > 0, inds$eggage + DELTAT, 0)
      tefi_out    <- tefi(inds$age, temp, food_eff, parms)
      inds$size   <- tefi_out$L
      neweggs     <- round(tefi_out$E)
      inds$eggs   <- ifelse(inds$size > SAM & inds$eggage == 0, pmax(0, neweggs), inds$eggs)
      inds
    })
  },

  #2 SURVIVE — GUTS lethal  + starvation proxy + maximum age
  survive = function(inds, parms) {
    with(parms, {
      N <- nrow(inds)
      if (N == 0) return(inds)

      food_dd  <- food_max / (1 + N / Nhalf)
      s        <- cT * pmax(0, inds$Dw - NEC)
      food_eff <- food_dd / (1 + s)
      e_scaled <- food_eff / food_max                   

      # GUTS stochastic death (hazard from damage above threshold zw)
      hz <- bw * pmax(0, inds$Dw - zw)
      ps <- exp(-hz * DELTAT)                             

      # Starvation proxy 
      p_starv <- (1 - e_scaled) * mortconst * DELTAT

      rnd_guts  <- runif(N)
      rnd_starv <- runif(N)
      
      # Survival
      subset(inds, rnd_guts < ps & rnd_starv > p_starv & inds$age < maxage)
    })
  },

  #3 HATCH (newborns inherit Dw = 0 via newdaphnia)
  hatch = function(inds, parms) {
    with(parms, {
      edt <- bottrell(temp)
      have.neo <- which(inds$eggs > 0 & inds$eggage > edt)
      eggs <- inds$eggs[have.neo]
      new.neo <- sum(eggs)
      inds$eggs[have.neo] <- inds$eggage[have.neo] <- 0
      rbind(inds, newdaphnia(L_0, new.neo))
    })
  }
)
```

---

# The solver  {#solver}

```{r solver}
myiteration <- function(y, ...) {
  observer <- function(res) {
    c(number   = nrow(res),
      meansize = mean(res$size),
      meanage  = mean(res$age),
      meaneggs = mean(res$eggs),
      meanDw   = mean(res$Dw))          # (added) population-mean scaled damage
  }
  init <- y@init; times <- fromtoby(y@times); func <- y@main
  parms <- y@parms; equations <- addtoenv(y@equations)
  environment(func) <- environment()
  parms$DELTAT <- 0
  out <- observer(init)
  for (i in 2:length(times)) {
    parms$DELTAT <- times[i] - times[i - 1]
    init <- func(times[i], init, parms)
    out <- rbind(out, observer(init))
  }
  row.names(out) <- NULL
  as.data.frame(cbind(times = times, out))
}
```

---

# Population scenarios {#run}

```{r run-scenarios}

run_rep <- function(model, Cw_value, seed) {
  parms(model)["Cw"] <- Cw_value
  set.seed(seed)
  out(sim(model))$number
}

scenarios <- list(
  "Control"     = 0,    # clean control
  "50 ug Cu/L"  = 25,   # ~ measured dissolved Cu; above NEC but below the lethal threshold
  "200 ug Cu/L" = 40    # ~ measured dissolved Cu; approaches the GUTS threshold -> lethal effects
)

n_rep <- 3                                      
days  <- seq(0, 75, by = 1)                   

envelopes <- lapply(scenarios, function(Cw) {
  reps <- sapply(1:n_rep, function(r) run_rep(ibm_cu, Cw, seed = 100 + r))
  data.frame(day = days,
             lo  = apply(reps, 1, min),
             hi  = apply(reps, 1, max),
             mid = apply(reps, 1, median))
})
```

```{r scenario-plot}
cols <- c("Control" = "black", "50 ug Cu/L" = "blue", "200 ug Cu/L" = "red")
par(mfrow = c(1, 2))
ymax <- max(sapply(envelopes, function(e) max(e$hi)), 100)
plot(NULL, xlim = c(0, 75), ylim = c(0, ymax),
     xlab = "Time (days)", ylab = "Total abundance",
     main = "Copper impacts on Daphnia")
for (nm in names(envelopes)) {
  e <- envelopes[[nm]]
  polygon(c(e$day, rev(e$day)), c(e$lo, rev(e$hi)), col = adjustcolor(cols[nm], 0.2), border = NA)
  lines(e$day, e$mid, col = cols[nm], lwd = 2)
}
legend("topleft", legend = names(scenarios), col = cols, lwd = 2, bty = "n")

# same plot, log scale
plot(NULL, xlim = c(0, 75), ylim = c(1, ymax), log = "y",
     xlab = "Time (days)", ylab = "Total abundance (log)", main = "")
for (nm in names(envelopes)) lines(envelopes[[nm]]$day, pmax(envelopes[[nm]]$mid, 1),
                                   col = cols[nm], lwd = 2)
```

---

# Validation

```{r read-pop-data}
pop <- read_excel("SI_population_data.xlsx")

pop$Day                  <- suppressWarnings(as.numeric(pop$Day))
pop$Nom_Cu_conc          <- suppressWarnings(as.numeric(pop$Nom_Cu_conc))
pop$Nom_Zi_conc          <- suppressWarnings(as.numeric(pop$Nom_Zi_conc))
pop$Average_no_organisms <- suppressWarnings(as.numeric(pop$Average_no_organisms))
pop <- pop[!is.na(pop$Average_no_organisms), ]

tot <- subset(pop, Type_of_organism == "Total" & Nom_Zi_conc == 0)
obs <- aggregate(Average_no_organisms ~ Nom_Cu_conc + Day, data = tot, FUN = mean)
nom2name <- c("0" = "Control", "50" = "50 ug Cu/L", "200" = "200 ug Cu/L") 
```

```{r validation plot}
par(mfrow = c(1, 3))

ymax_v <- max(c(obs$Average_no_organisms,
                sapply(envelopes, function(e) max(e$hi))), na.rm = TRUE)
for (nom in c(0, 50, 200)) {
  nm <- nom2name[as.character(nom)]; e <- envelopes[[nm]]; o <- subset(obs, Nom_Cu_conc == nom)
  plot(o$Day, o$Average_no_organisms, pch = 15, col = "darkblue",
       xlim = c(0, 75), ylim = c(0, ymax_v), xlab = "Time (days)", ylab = "Total abundance",
       main = paste0(nom, " ug/L"))
  polygon(c(e$day, rev(e$day)), c(e$lo, rev(e$hi)), col = adjustcolor("grey40", 0.5), border = NA)
  lines(e$day, e$mid, col = "black", lwd = 2)
  points(o$Day, o$Average_no_organisms, pch = 15, col = "darkblue")   
  legend("topright", legend = c("Observed", "Predicted"),
         pch = c(15, NA), lwd = c(NA, 2), col = c("darkblue", "black"), bty = "n", cex = 0.9)
}
```

```{r rsq}
# log-abundance R2: compare log(observed) with log(predicted median)
rsq_log <- function(nom) {
  nm <- nom2name[[as.character(nom)]]
  e  <- envelopes[[nm]]
  o  <- obs[obs$Nom_Cu_conc == nom, c("Day", "Average_no_organisms")]
  o  <- o[is.finite(o$Average_no_organisms) & o$Average_no_organisms > 0, ]
  pred <- e$mid[match(o$Day, e$day)]           
  keep <- is.finite(pred) & pred > 0
  if (sum(keep) < 2) return(NA_real_)           
  yo <- log(o$Average_no_organisms[keep]); yp <- log(pred[keep])
  1 - sum((yo - yp)^2) / sum((yo - mean(yo))^2)
}
data.frame(Treatment = c("Control", "50 ug Cu/L", "200 ug Cu/L"),
           R2_log = round(sapply(c(0, 50, 200), rsq_log), 2))
```


```{r damage-diagnostic}
set.seed(7)
dmg <- sapply(scenarios, function(Cw) { parms(ibm_cu)["Cw"] <- Cw
out(sim(ibm_cu))$meanDw })
matplot(days, dmg, type = "l", lty = 1, lwd = 2, col = cols,
        xlab = "Time (days)", ylab = "Mean scaled damage Dw (µg/L)",
        main = "Damage build-up vs effect thresholds", ylim = c(0, 45))
abline(h = 12.35, lty = 3); text(62, 14, "NEC (sublethal)", cex = 0.8)
abline(h = 38.24, lty = 2); text(62, 40, "zw (lethal)",     cex = 0.8)
legend("bottomright", legend = names(scenarios), col = cols, lwd = 2, bty = "n")
```

