---
title: "Preprocessing for ecotoxicological prediction in R"
output:
  html_document:
    df_print: paged
---

## Introduction

In this session we aim to link ecotoxicological data to databases that provide data for prediction and to exemplarily compute the similarity between chemicals.

To work with realistic data, we will use data from the EPA Ecotoxicology database. We use the data from the `standartox` R package, which has undergone some pre-processing and outlier removal.

```{r, echo = FALSE}
setwd("/Users/ralfs/Arbeit/Lehre_Betreuung/2025/Theoretical_ecotox/Lecture_9")
library(standartox)
library(stringr)
library(dplyr)
library(chemlook)
# no CRAN package - is installed via:
# remotes::install_github('andschar/chemlook')
library(webchem)
library(rcdk)
library(fingerprint)
library(ECOTOXr)
library(biomonitoR)
library(taxize)
library(R.matlab)
# to install biomonitoR
# library(devtools)
# install_github("alexology/biomonitoR", ref = "main", build_vignettes = TRUE)
```

```{r, eval = FALSE}
# Query database and store locally to avoid querying each time
dat <- stx_data(dir_out = "/Users/ralfs/R/0_datasets/standartox")
tax <- stx_taxa(dir_out = "/Users/ralfs/R/0_datasets/standartox")
che <- stx_chem(dir_out = "/Users/ralfs/R/0_datasets/standartox")
# merge data
dat2 <- merge(dat, tax, by = 'tl_id', all.x = TRUE)
dat3 <- merge(dat2, che, by = 'cl_id', all.x = TRUE)
# save(dat3, "Standartox.Rdata")
# to load requires the bit64 package
```

```{r, echo = FALSE}
# only execute when you have saved the .Rdata file
library(bit64) # needed to interpret some formats
load('Standartox.Rdata')
```

We limit the dataset to Ephemeropterans, measured toxicity (i.e. ignoring *larger than* values), mortality, median toxic concentrations *XX50* and to test durations between 24 and 120 h. Then we check the taxonomic level of the results as well as the units of the concentrations.

```{r}
dat4 <- dat3 %>% filter(qualifier == "=" & order == "Ephemeroptera" & effect == "mortality" & endpoint_group == "XX50" & between(duration, 24, 120))

# display concentration units and level
dat4 %>% select(concentration_unit, rank) %>% table()
```

A relevant fraction of tests has been done with organisms that were only identified to family or genus level. It remains open whether the tests were done with mixtures of taxa or identification was limited to higher taxonomic levels. The most likely explanation is that the taxa were not identified to species level. How to deal with this situation depends on the purpose of the study. If the aim is, for example, prediction and the data would be considered independent, this could incur bias if in reality the same taxon was tested. Here, we limit the data to the species level.

```{r}
dat4_spec <- dat4[rank == "species", ]
```

Moreover, we bring all concentrations to the same unit. As an approximation, ppb corresponds to µg L^-1^. We create a small function for conversion and then convert the concentrations where the unit of the concentration is ppb. This can easily be done with the `case_when()` function used within `mutate()`.

```{r}
# convert ppb to g/l
ppb_to_gL <- function(ppb) {
  ppb * 1e-6
}


dat4_fin <- dat4_spec %>%
  mutate(concentration = case_when(
  			concentration_unit == "ppb" ~ ppb_to_gL(concentration),
      	TRUE ~ concentration), 
      	concentration_unit = "g/L")
```

The dataset contains multiple entries per species-chemical combination, which we aggregated with the geometric mean.

```{r}
dat5 <- dat4_fin %>%
			group_by(family, genus, taxon, casnr, cas, cname, concentration_unit) %>%
  			summarise(geo_mean = exp(mean(log(concentration))), .groups = "drop")

```

We now identify the unique chemicals in the data set for which we extract physicochemical information and compute the similarity.

```{r}
chemicals <- dat5 %>% 
					select(casnr, cas, cname) %>% 
  					distinct()
# count chemicals
nrow(chemicals)
```

We have 232 different chemicals. We will only focus on a subset, because most queries will connect to servers and this would otherwise take a bit longer. First, we use the function `websearch_comptox()` of the `ECOTOXr` package to connect to the [COMPTOX dashboard](https://comptox.epa.gov/dashboard) of the US EPA, which provides a wide range of information on chemicals and their toxicity.

```{r}
# subset to first 10 chemicals
chem_sub <- chemicals[1:10, ] 

# query comptox
comptox_results <- websearch_comptox(chem_sub$cas)

# look at results
comptox_results
```

We obtain a wide range of data.

```{r}
comptox_results$`Main Data`

```

**Inspect the data yourself. Why do some compounds consistently lack information for the octanol water partition coefficient, water solubility, vapor pressure and similar variables?**

The package `ECOTOXr` provides also a function to access the Ecotoxicology database of the US EPA, i.e. the raw data that has been polished in the Standartox database, but this is beyond the scope here and described in an article by [Vries (2024)](https://doi.org/10.1016/j.chemosphere.2024.143078).

We can also query physico-chemical properties from the [PubChem database](https://pubchem.ncbi.nlm.nih.gov). This can be done using the function `pc_prop` in the package `webchem`. For detailed information on webchem see the article by [Szöcs et al. (2020)](https://www.jstatsoft.org/article/view/v093i13). To use this function, we first need the so-called Pubchem Compound ID (CID), which can be queried using the function `get_cid()`.

```{r}
cid_demo <- get_cid(chem_sub$cas, from = "cas")
cid_demo
```

We obtain an object with 30 rows. **Why 30 rows, can you explain what happened?**

We continue querying physico-chemical properties for the retrieved CIDs.

```{r}
props_demo <- pc_prop(cid_demo$cid, properties = c("Title", "MolecularWeight", "XLogP"))
props_demo
```

We notice again that for several compounds the Log*P* is not defined, this is not surprising because most of these are inorganic chemicals.

Let us take a step back: We have seen that a CAS number can be associated with multiple CIDs in PubChem. To obtain unambiguous results, we can use the Inchi key:

```{r}
# we obtained the Inchi key when we queried the Comptox database
inchi_sample <- comptox_results$`Main Data`$INCHIKEY

cid_demo_2 <- get_cid(inchi_sample, from = "inchikey")
cid_demo_2
```

Even with this, we obtain multiple CIDs per Inchi key. This is more a database issue and requires manual checking. For example, if you would use a large language model or manually query PubChem, you would see that the CIDs 518533, 12820921 and 21924987, which all relate to the Inchi key NLXLAEXVIDQMFP-UHFFFAOYSA-N, all represent ammonium chloride. PubChem often has several distinct CIDs for closely related ionic species, salts, or registry duplicates. All three relate to ammonium/chloroaminium chemistry but represent slightly different forms or records. One way to obtain unique CIDs is to query a [database curated by Andreas Scharmüller](https://github.com/andschar/chemlook) that is provided in the package `chemlook`.

```{r}
chem_info <- cl_query(chem_sub$cas, from = "cas")

```

At least we obtained 5 unique CIDs and could proceed with the Inchi key for the missing ones to have complete information.

**Try to obtain unique CIDs for the missing compounds using different approaches (e.g. querying CIDs with SMILES, manual checking).**

Let us use now compute the Tanimoto similarity. This can easily be done with the packages `rcdk` and `fingerprint` and requires the SMILES codes. We import the SMILES code into the CKD format with the function `parse.smiles`.

```{r}
mols <- parse.smiles(comptox_results$`Main Data`$SMILES)
fps <- lapply(mols, get.fingerprint, type = "extended")
```

Different fingerprints are available, use the help for the function `get.fingerprint()` for an overview. Now that we have fingerprints, we compute the similarity

```{r}
fp.sim <- fingerprint::fp.sim.matrix(fps, method='tanimoto')
fp.sim
```

Similarities are generally low. We can cluster these for visualisation.

```{r}
# convert similarity to distance
fp.dist <- 1 - fp.sim
cls <- hclust(as.dist(fp.dist))
# shorten labels
labels <- abbreviate(comptox_results$`Main Data`$PREFERRED_NAME, minlength = 9)
labels
# plot
plot(cls, main = "", labels = labels)
```

Not surprisingly, ammonia and ammonium chloride have the highest similarity. In the package [vignette](https://cran.r-project.org/web/packages/rcdk/vignettes/using-rcdk.html) it is also described how you could identify the most similar target molecule, for example if you wanted to use this in prediction. In addition, the vignette further shows what you can do with the `rcdk` package.

If you are interested, also checkout the package `ChemmineR`, which provides multiple functions for chemical informatics. Consider going through the [online tutorial](https://www.bioconductor.org/packages/devel/bioc/vignettes/ChemmineR/inst/doc/ChemmineR.html#Searching_PubChem).

Finally, we check out biological and ecological complementary information that may be useful for prediction. As for the chemicals, we first extract the unique taxa we have.

```{r}
ourtaxa <- dat5 %>% 
					select(family, genus, taxon) %>% 
  					distinct()
# count chemicals
nrow(ourtaxa)
ourtaxa
```

We have a range of taxa from Ephemeroptera. For several analyses, it can be relevant to have the full taxonomic tree and check whether the names are spelt correctly. This can be done with the package `taxize`. We first resolve the names.

```{r}
temp_names <- gna_verifier(ourtaxa$taxon)
temp_names
```

We obtain a whole bunch of information including a cardinality score, which tells us something about the degree of uniqueness or ambiguity of our query name in the database. The score ranges from 1 to 0, where 1 means that the name uniquely and confidently matches one taxon. The variable *acceptedNameScore* is a bit simpler and tells us whether the name is accepted (valid) in the database. This is often sufficient. Let us quickly check the matched database names and scores.

```{r}

temp_names %>% select(submittedName, matchedName, acceptedNameScore)

```

All fine, we do not have to deal with spelling errors or alike. We can also use the `taxize` package to add taxonomic information. This is done with the function `classification()`. This takes a bit, therefore we limit this to the first 10 taxa.

```{r, echo=FALSE}
taxaclass <- classification(ourtaxa$taxon[1:10], db = 'ncbi')
# 8 of 10 found
taxaclass2 <- classification(ourtaxa$taxon[1:10], db = 'gbif')
# all taxa are found
taxaclass2
```

We obtain information on the taxonomy from kingdom to species. We can also query specific information, e.g. family level:

```{r, echo=FALSE}
family_temp <-tax_name(ourtaxa$taxon[1:10], get = "family", db = "both")
# 9 of 10 found
```

Except for *Caenis maxima*, we have found the families for all taxa. We could also query all downstream taxa. For example, if we only had family information, to get an idea of how many taxa exist in this taxonomic branch. Let us check this for the taxa that were tested at family level.

```{r, echo = FALSE}
dat4_fam <- dat4[rank == "family", ]
# extract unique families
dat_families <- dat4_fam %>%
                select(family)  %>%
                distinct()

# Watch out - running the following function can take quite a bit of time.
down_families <- downstream(dat_families$family, downto = "species", db = "gbif")
head(down_families[[1]])
head(down_families[[2]])
```

This yields to 250 taxa, though the list may still have some gaps.

Trait information on taxa is available in multiple databases, but direct access through R is patchy and not straightforward. Examples include the packages:

-   [traits](https://docs.ropensci.org/traits/), which covers a range of organism groups
-   [TR8](https://cran.r-project.org/web/packages/TR8/index.html), which covers different databases for plants
-   [rtry](https://cran.r-project.org/web/packages/rtry/index.html), which simplifies working with the TRY plant trait database
-   [biomonitoR](https://github.com/alexology/biomonitoR), which provides some traits for freshwater organisms.

Theoretically, the `biomonitoR` package could be used to add traits of freshwater organisms to our data, but it relies on an internal reference database for taxa, which makes its use a bit cumbersome if our taxa are not included. However, let's use it to add some trait information. We can use the function `assign_traits()` to assign trait information from the [Tachet database](https://doi.org/10.1023/A:1017042921298).

```{r}
ourtaxa_up <- ourtaxa %>%
              mutate(Taxa = taxon, 
                     Site = 10)
# function requires a column called Taxa and abundances in a site
# as a workaround, we add a dummy column with abundances
data_bio <- as_biomonitor(ourtaxa_up[ , 4:5])
data_bio
# most taxa not found (as given under suggested taxon names)

# but we can use an own tree to improve the matching
dfref_gbif <- get_gbif_taxa_tree(ourtaxa_up$Taxa)
# a few taxa are synonyms of accepted names
dfref_gbif$synonym
# replace synonym in our data by accepted name
ourtaxa_up2 <- ourtaxa_up %>%
  left_join(dfref_gbif$synonym, by = c("Taxa" = "synonym")) %>%
  mutate(Taxa = if_else(!is.na(accepted), accepted, Taxa)) %>%
  select(-accepted)
ourtaxa_up2

# now we can convert to a biomonitor object
data_asb <- as_biomonitor(ourtaxa_up2[ , 4:5], dfref = dfref_gbif$taxonomy)
data_asb
# one taxon was removed because it had no match
# we need to run the function aggregate_taxa before, otherwise trait assignment does not work.
# This function harmonizes taxonomic levels

data_asb_agg <- aggregate_taxa(data_asb)
data_asb_agg

# Now we assign traits
data_ts <- assign_traits(data_asb_agg, tax_lev = "Family")
data_ts
```

Now we have some traits at the family level for our species. However, the function would not run at genus level because of too few matches. Generally, it may be easier and more flexible to import the trait database into R and link the data manually. For example, for freshwater organisms, a more current database is the freshwaterecology.info database, which can be [accessed via API](https://www.freshwaterecology.info/fweapi2docu.php#docur). For further details on using the biomonitoR package, in particular for calculating ecological metrics, have a look at the vignette. 
```{r, eval = FALSE}
vignette("functional", package = "biomonitoR")
```

Moreover, a recent paper on this package by [Laini et al. 2022](https://peerj.com/articles/14183/) provides commented R code in the Supplemental information that may be helpful.

Finally, to access physiological information for species, we could query the [Add my pet database](https://www.bio.vu.nl/thb/deb/deblab/add_my_pet/). There, we can download the .zip file *AmPdata*, which contains data in the Matlab format. With the package `R.matlab` we can read these files.

```{r, eval = FALSE}
# NOT run
# you would need to download the data into your working directory path, but the data set is relatively big and takes several minutes to import.

# Load the mat files
allStat <- readMat("AmPdata/allStat.mat")
popStat <- readMat("AmPdata/popStat.mat")
allUnits <- readMat("AmPdata/allUnits.mat")
allLabel <- readMat("AmPdata/allLabel.mat")
```

```{r}
# But we can load an R object where I saved the information
# Will not run on your machine as long as you do not change the path
load('~/Arbeit/Lehre_Betreuung/2025/Theoretical_ecotox/Lecture_9/allStat_addmypet.RData')
```

The loaded objects are complex, often nested lists. For example:
```{r}
str(allStat[[1]][1])
```

Each of the list corresponds to a species. For example, if we searched for data on Daphnia magna, we could extract this as follows.
```{r}
# we need to set up a function that searches the lists
get_amp_species <- function(species_name, data_file) {
  for (entry in data_file[[1]]) {
    if (!is.null(entry[[1]]) && entry[[1]][1] == species_name) {
      return(entry)
    }
  }
  return(NULL)
}


speciesName <- "Daphnia_magna"
# note that the names are saved with an underscore
speciesData <- get_amp_species(speciesName, allStat)
speciesData
```
From there, you could convert it into a data.frame or tibble for analysis.

As a final step, let us have a look, which species, families and orders are included.
```{r}
species_names <- sapply(allStat[[1]], function(x) x[[1]][1])
head(species_names, 10)
# note that this would still contain duplicates

# extract families
family_names <- sapply(allStat[[1]], function(x) x[[3]][1])
# we remove duplicates
famil_nam_dedup <- unique(family_names)
# only 1041 of 6012 left

order_names <- sapply(allStat[[1]], function(x) x[[4]][1])

# extract all families of order Ephemeroptera
families_ephem <- family_names[order_names == "Ephemeroptera"]
families_ephem
```

Only 4 families are represented in the database, which relate to 7 taxa.

**If you want to practice, choose a different organism group in the beginning, which would also translate into a different set of tested chemicals, and re-run the code for this!**





